---
title: "Electric Vehicles Final Project Stat184"
author: "Owen McInerney, Eric Best, Anirudh Kumar"
date: "2023-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r load_packages, include = FALSE}
# Style of Choice -- TidyVerse Styling
library(groundhog)
groundhog.day <- "2023-11-20"
pkgs <- c("here", "dplyr", "dbplyr", "ggplot2", "readxl", "tidyverse", "knitr", "kableExtra", "leaflet")
groundhog.library(pkgs, groundhog.day)
```
## Introduction

## Research Questions

(1) How common are charging stations in Pennsylvania for public use? 
(2) What vehicle has the longest excursion time/length on a full-charge?
(3) How do EV manufacturers compare in excursion range and vehicle price?
(4) How has global EV usage changed over the years?
(5) Which countries lead in EV usage?  Which countries are lagging behind?
(6) How do Electric Vehicles and Gas Vehicles compare in MPGe?

## Data Sources

-
-
-
-

## Data Wrangling

``` {r load_data}
options(scipen = 999) # Ensures scientific notation is not used

# IEA csv File for Q4 and Q5
csv_path <- here("IEA_EV_SalesHistoricalCars.csv")
EV_Global_Historical_raw <- read.csv(csv_path)

# US department of Energy xlsx File for Q6
EV_GAS_data_raw <- read_excel("all_alpha_24.xlsx")

# BLS Data on Fuel Prices xlsx File for Q7
gas_prices <- read_excel("bls_gas_prices.xlsx", range = "B11:M11", col_names = FALSE)
electricty_prices <- read_excel("bls_electricity_prices.xlsx", range = "B11:M11", col_names = FALSE)
```

``` {r wrangle_data}
### Wrangle Data ---------------------------------------------------------------
## Q4: wrangle IEA data to cases of years with attributes: year, count of EVs
EV_supply_global_year <- EV_Global_Historical_raw %>%
  filter(unit == "Vehicles", region != "World") %>%
  select(year, value) %>%
  group_by(year) %>%
  summarise(count = sum(value))

## Q5: wrangle IEA data to get cases of regions from 2019-2022 with attributes: region, count of EVs
EV_supply_region_2019_2022 <- EV_Global_Historical_raw %>%
  filter(unit == "Vehicles", region != "World", year >= 2019) %>%
  select(region, value) %>%
  group_by(region) %>%
  summarise(count = sum(value))

## Q6: US department of energy data: only want to include Model, fuel, veh_class, air pollution score, city mpg, highway mpg,  greenhouse gas score

# Change names of columns with spaces
colnames(EV_GAS_data_raw)[11] <- "veh_class"
colnames(EV_GAS_data_raw)[12] <- "air_pollution_score"
colnames(EV_GAS_data_raw)[13] <- "city_mpge"
colnames(EV_GAS_data_raw)[14] <- "highway_mpge"
colnames(EV_GAS_data_raw)[15] <- "cmb_mpge"
colnames(EV_GAS_data_raw)[16] <- "greenhouse_gas_score"

# Cases are cars with attributes: Model, fuel, vehicle class, air pollution score, city mpg, highway mpg, greenhouse gas score
EV_GAS_data <- EV_GAS_data_raw %>%
  drop_na(city_mpge) %>%
  filter(Fuel == "Gasoline" | Fuel == "Electricity") %>%
  select(Model, Fuel, veh_class, air_pollution_score, city_mpge, highway_mpge, cmb_mpge, greenhouse_gas_score) %>%
  mutate(
    city_mpge = as.numeric(city_mpge),
    highway_mpge = as.numeric(highway_mpge),
    cmb_mpge = as.numeric(cmb_mpge),
    greenhouse_gas_score = as.numeric(greenhouse_gas_score)
  )

## Q7: BLS Data + US department of energy data: case = a month with attribute: month, miles per gallon equivalent for fuel and electricity
electricty_prices[1:12] = electricty_prices[1:12] * 33.7 # converts from kWh to kWh equivalent of a gallon
fuel_df = full_join(gas_prices, electricty_prices)

# Currently have each month as a given column: need to label columns
colnames(fuel_df)[1] <- "jan"
colnames(fuel_df)[2] <- "feb"
colnames(fuel_df)[3] <- "mar"
colnames(fuel_df)[4] <- "apr"
colnames(fuel_df)[5] <- "may"
colnames(fuel_df)[6] <- "jun"
colnames(fuel_df)[7] <- "jul"
colnames(fuel_df)[8] <- "aug"
colnames(fuel_df)[9] <- "sep"
colnames(fuel_df)[10] <- "oct"
colnames(fuel_df)[11] <- "nov"
colnames(fuel_df)[12] <- "dec"

# To find average mpge for gas and electric
mpge_avg_fuel <- EV_GAS_data %>%
  group_by(Fuel) %>%
  summarise(mpge_avg = mean(cmb_mpge))

electricity_price_per_mile <- mpge_avg_fuel %>%
  filter(Fuel == "Electricity") %>%
  select(mpge_avg)

gas_price_per_mile <- mpge_avg_fuel %>%
  filter(Fuel == "Gasoline") %>%
  select(mpge_avg)

# Make sure they are numeric
electricity_avg_mpge <- as.numeric(electricity_price_per_mile)
gas_avg_mpge <- as.numeric(gas_price_per_mile)

# Add fuel type column to data frame
fuel_df$type = c("gas", "electricity") 

# Need to change cases to months 
fuel_df <- fuel_df %>%
  pivot_longer(cols = c(jan, feb, mar, apr, may, jun, jul, aug, sep, oct, nov, dec), 
               names_to = "month", 
               values_to = "avg_gallon_equaivalent_price")

# Making actual dates instead of character dates
dates <- append(seq(from = as.Date("2022-01-01"), to = as.Date("2022-12-01"), by = 'month'), 
                seq(from = as.Date("2022-01-01"), to = as.Date("2022-12-01"), by = 'month'))

# Add dates column to fuel_df
fuel_df$date = dates

# do not need month column anymore, want to mutate from avg price per gallon to miles per dollar
fuel_df <- fuel_df %>%
  select(type, avg_gallon_equaivalent_price, date) %>%
  mutate(miles_per_dollar = case_when(type == 'gas' ~ gas_avg_mpge / avg_gallon_equaivalent_price,
         type == 'electricity' ~ electricity_avg_mpge / avg_gallon_equaivalent_price)) %>%
  select(date, type, miles_per_dollar)
```

## Question 4: How has the global supply of electric vehicles changed over the years?

Around 1900, steam vehicles, gas vehicles, and electric vehicles were all competitive with each other.  Gas vehicles were more difficult to drive, while electric vehicles were easier to drive.  With electricity becoming more popular in homes, electric vehicles became a viable option for short trips.  However, when Henry Ford came out with the Model T, which cost only $650 per car, electric cars could not compete at $1750 per car.  The world became accustomed to gas-powered vehicles, and gas still is the most common fuel source for vehicles.  As science developed, electric cars have became cheaper and more efficient. In addition climate change, partially induced by gas-powered vehicles, has been a reason to make the switch to electric vehicles.  Another reason for the switch to electric vehicles, in sudden gas shortages caused by war and dependence on other countries. 

Our group wanted to investigate how global EV supply has changed over the year.  Using the 2023 Global EV Outlook data set from the International Energy Agency, we were able to examine the change in the supply of electric vehicles over the years.  From the bar chart below, we can see that the quantity of electric vehicles' supplied has increased exponentially each year since 2010.  We can predict from this that there will be another significant increase in electric vehicles supplied for 2023.
```{r q4_bar_plot}
### Create Bar Chart of Global EV Ownership by Year -----------------------------
global_ev_supply_by_year_bar_chart <- ggplot(EV_supply_global_year, aes(x = year, y = count, fill = "red")) +
  geom_col() +
  scale_y_continuous(n.breaks = 8, limits = c(0, max(EV_supply_global_year$count))) +
  scale_x_continuous(n.breaks = 11) +
  labs(
    title = "Global Electric Vehicle Supply since 2010",
    x = "Year",
    y = "Quantity of Electric Vehicles Supplied"
  ) +
  theme_classic() +
  theme(legend.position = "none") # removes legend

## View Global EV supply by year Bar Chart
global_ev_supply_by_year_bar_chart
```

## Question 5: Which countries lead in EV usage?  Which countries are lagging behind?

Similar to the race to the moon, different regions are trying to adapt quickly to the need for electric vehicles.  Our group wanted to investigate which countries lead in the race to the adaption to electric vehicles.  We also wanted to see which regions implemented electric vehicles the least. 

From the bar chart below, we can see that China have the greatest quantity of electric vehicles supplied.  This might be due to the fact that they are the most populated region in the world.  It also may be due to their lack of child labor laws, which would allow them to increase production at a smaller cost.  We can also see that Turkey, South Africa, and Chile have implemented electric cars the least.  This is likely due to their economies not being as strong as China's or the United States.

```{r q5_bar_plot}
## Create bar chart of EV Ownership by Country ---------------------------------
# use y = reorder(region, +count) to order countries by car ownership
regional_ev_supply_bar_chart <- ggplot(EV_supply_region_2019_2022, aes(x = count, y = reorder(region, +count), fill = "red")) +
  geom_col() +
  scale_x_continuous(limits = c(0, max(EV_supply_region_2019_2022$count)), n.breaks = 6) +
  labs(
    title = "Electric Vehicle Supply By Region",
    subtitle = "Total Electric Vehicles Supplied from 2019 to 2022 for each Region",
    x = "Quantity Supplied of Electric Vehicles",
    y = "Region"
  ) +
  theme_classic() +
  theme(legend.position = "none") # removes legend

## View Regional EV Supply Bar Chart
regional_ev_supply_bar_chart
```

## Question 6: How do Electric Vehicles and Gas Vehicles compare in MPGe?

Our group wanted to investigate whether gas-powered vehicles or electric vehicles were more efficient.  We decided to make a box plot of the miles per gallon (mpge) equivalent (where 33.7 kWh = 1 gallon) for gas-powered vehicles and electric vehicles.

In the box plot below, we can see that electric vehicles tend to have a higher mpge than gas-powered vehicles.  This may be due to investment in electric vehicles which has allowed for quick advancement in the range of electric vehicles.  Also, electric vehicles directly convert a kilowatt hour to motion while gas-powered vehicle convert gas into heat which then creates motion.

```{r q6_box_plot}
### Make Box Plots -------------------------------------------------------------
## Combined MPGe Box Plot
combined_mpge_box_plot <- EV_GAS_data %>%
  group_by(Fuel) %>%
  ggplot(aes(x = Fuel, y = cmb_mpge, group = Fuel, fill = Fuel)) +
  labs(title = "City and Highway MPGe EV vs Gas Box Plot", subtitle = "MPGe is Miles per Gallon equivalent", x = "Fuel", y = "MPGe") +
  scale_y_continuous(n.breaks = 10, limits = c(0, max(EV_GAS_data$cmb_mpge))) +
  geom_boxplot(
    outlier.colour = "black", outlier.shape = 16,
    outlier.size = 2, notch = FALSE
  ) +
  theme_minimal()

## View Combined MPGe Box Plot
combined_mpge_box_plot
```

## Question 7: How do costs differ between electric vehicles and gas vehicles?

After seeing that electric vehicles tend to have a greater miles per gallon equivalent than gas-powered vehicles, we wanted to see which was more cost effective.  We decided to take date from the Bureau of Labor Statistics for energy costs each month during 2022 (because 2023 has not finished yet).  We also took the average mpge from the 2024 vehicles tested by the US department of energy.  We then divided the mpge by the energy costs for each month and fuel type.  

We can see from the line graph below that an electric vehicle will go further distance per dollar of energy than a gas-powered vehicle.  This is due to electric vehicles having nearly four times the average mpge than a gas-powered vehicle.  Future evaluations should include costs to purchase the vehicles and costs of maintaining the vehicles.
```{r q7_line_graph}
### Make Line Graph ------------------------------------------------------------
Miles_per_dollar_line_graph <- fuel_df %>%
  ggplot(aes(x = date, y = miles_per_dollar, color = type)) +
  geom_line() +
  scale_y_continuous(n.breaks = 10, limits = c(0, 20)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(
    title = "Miles per Dollar by Month: Electricity Vs Gas",
    subtitle = "Using 2024 vehicles and 2022 monthly fuel prices",
    x = "Month", y = "Miles per Dollar", color = "Fuel Type"
  ) +
  theme_bw()

## View Miles per Dollar Line Graph
Miles_per_dollar_line_graph
```
